### 2016-09-08 paper/icfp16-cameraready/abstract.tex

\begin{abstract}
  %
  \emph{Static} type errors are a common stumbling block
  for newcomers to typed functional languages.
  %
  We present a \emph{dynamic} approach to explaining type
  errors by generating counterexample witness inputs that
  illustrate \emph{how} an ill-typed program goes wrong.
  %
  First, given an ill-typed function, we symbolically
  execute the body to dynamically synthesize witness values
  that can make the program go wrong.
  We prove that our procedure synthesizes
  \emph{general witnesses} in that if a witness is
  found, then \emph{for all} inhabited input types, there exist
  values that can make the function go wrong.
  %
  Second, we show how to extend the above procedure to
  produce a \emph{reduction graph} that can be used to
  interactively visualize and debug witness executions.
  %
  Third, we evaluate our approach on two data sets
  comprising over 4,500 ill-typed student programs.
  Our technique is able to generate witnesses for
  88\% of the programs, and our reduction graph
  yields small counterexamples for 81\% of the witnesses.
\end{abstract}

>>>>

\begin{abstract}
  %
  {Static} type errors are a common stumbling block
  for newcomers to typed functional languages.
  %
  We present a {dynamic} approach to explaining type
  errors by generating counterexample witness inputs that
  illustrate how an ill-typed program goes wrong.
  %
  First, given an ill-typed function, we symbolically
  execute the body to synthesize witness values that
  make the program go wrong.
  We prove that our procedure synthesizes
  {general} witnesses in that if a witness is
  found, then for all inhabited input types,
  there exist values that can make the function go wrong.
  %
  Second, we show how to extend the above procedure to
  produce a reduction graph that can be used to
  interactively visualize and debug witness executions.
  %
  Third, we evaluate the coverage of our approach
  on two data sets comprising over 4,500 ill-typed
  student programs.
  Our technique is able to generate witnesses for
  88\% of the programs, and our reduction graph
  yields small counterexamples for 81\% of the witnesses.
  %
  Finally, we evaluate whether our witnesses help
  students understand and fix type errors, and
  find that students presented with our witnesses
  show a greater understanding of type errors
  than those presented with a standard error message.
\end{abstract}

### 2016-09-08 paper/icfp16-cameraready/introduction.tex

Type errors are a common stumbling block for students
trying to learn typed functional languages like \ocaml\
and \haskell.
%
Consider the ill-typed @fac@ function on the left in
Figure~\ref{fig:factorial}.
%
The function returns @true@ in the base case (instead of @1@),
and so \ocaml responds with the error message:
%
\begin{verbatim}
 File "fac.ml", line 5, characters 8-17:
 Error: This expression has type bool but an
        expression was expected of type int.
\end{verbatim}
%
This message makes perfect sense to an expert who is familiar
with the language and has a good mental model of how the type
system works.
%
However, it may perplex a novice who has yet to develop such a
mental model.
%
To make matters worse, unification-based type inference algorithms
often report errors far removed from their source.
%
This further increases the novice's confusion and can actively mislead
them to focus their investigation on an irrelevant piece of code.
%
Much recent work has focused on analyzing unification constraints
to properly \emph{localize} a type error~\cite{Lerner2007-dt,Chen2014-gd,Zhang2014-lv,Pavlinovic2014-mr},
but an accurate source location does not explain \emph{why} the
program is wrong.

\begin{figure}[t]
\centering
\begin{minipage}{.49\linewidth}
\centering
\begin{code}
  let rec fac n =
    if n <= 0 then
      true
    else
      n * fac (n-1)
\end{code}
\vspace{3em}
% \end{minipage}
% % \hspace{0.3in}
% \begin{minipage}{.39\linewidth}
\centering
  \includegraphics[height=1.5in]{fac-overview.png}
\end{minipage}
\begin{minipage}{.49\linewidth}
\centering
\includegraphics[height=3in]{fac-long.png}
\end{minipage}
\caption{(top-left) An ill-typed \texttt{fac} function; (bottom-left) Dynamically witnessing the type error in \texttt{fac}, showing only function calls; (right) The same trace, fully expanded.}
\label{fig:factorial}
\end{figure}

>>>>

Type errors are a common stumbling block for students
trying to learn typed functional languages like \ocaml\
and \haskell.
%
Consider the ill-typed @fac@ function on the left in
Figure~\ref{fig:factorial}.
%
The function returns @true@ in the base case (instead of @1@),
and so \ocaml responds with the error message:
%
\begin{verbatim}
  This expression has type
    bool
  but an expression was expected of type
    int.
\end{verbatim}
%
This message makes perfect sense to an expert who is familiar
with the language and has a good mental model of how the type
system works.
%
However, it may perplex a novice who has yet to develop such a
mental model.
%
To make matters worse, unification-based type inference algorithms
often report errors far removed from their source.
%
This further increases the novice's confusion and can actively mislead
them to focus their investigation on an irrelevant piece of code.
%
Much recent work has focused on analyzing unification constraints
to properly \emph{localize} a type error~\cite{Lerner2007-dt,Chen2014-gd,Zhang2014-lv,Pavlinovic2014-mr},
but an accurate source location does not explain \emph{why} the
program is wrong.

\begin{figure}[t]
\centering
\begin{minipage}{.49\linewidth}
\centering
\begin{ecode}
let rec fac n =
  if n <= 0 then
    true
  else
    n * __fac (n-1)__
\end{ecode}
\vspace{3em}
% \end{minipage}
% % \hspace{0.3in}
% \begin{minipage}{.39\linewidth}
\centering
  \includegraphics[height=1.5in]{fac-overview.png}
\end{minipage}
\begin{minipage}{.49\linewidth}
\centering
\includegraphics[height=3in]{fac-long.png}
\end{minipage}
\caption{(top-left) An ill-typed \texttt{fac} function, the error location reported by \ocaml is underlined; (bottom-left) Dynamically witnessing the type error in \texttt{fac}, showing only function calls; (right) The same trace, fully expanded.}
\label{fig:factorial}
\end{figure}

### 2016-09-08 paper/icfp16-cameraready/related.tex

A function's type is a theorem about the function's behavior.
Thus, \toolname's witnesses can be viewed as \emph{counterexamples},
thereby connecting it to work on using test generation to find
counterexamples prior to starting a proof proof~\cite{ACL2Testing, Seidel15}.

@@

\cite{Flanagan96} describes a static debugger for Scheme, which helps
the programmer interactively visualize problematic source-sink flows
corresponding soft-typing errors. The debugger allows the user to explore
an abstract reduction graph computed from a static value set analysis of
the program. In contrast, \toolname generates witnesses and allows the user
to explore the resulting dynamic execution.
%
Perera \etal~\cite{Perera2012-dy} present a tracing semantics
for functional programs that tags values with their provenance, enabling
a form of backwards program slicing from a final value to the sequence
of reductions that produced it. Notably, they allow the user to supply a
\emph{partial value} --- containing holes --- and present a partial slice,
containing only those steps that affected the the partial value.
% This
% system is designed to answer questions of the form ``Where did this
% value come from?'' and thus is focused on backward exploration.
Perera~\etal\ focus on backward exploration; in contrast, our
visualization supports forward \emph{and} backward exploration, though
our backward steps are more limited.
%
Specifically, we do not support selecting a value and inserting the
intermediate terms that preceded it while ignoring unrelated computation
steps. %; this would be interesting future work.

>>>>

A function's type is a theorem about the function's behavior.
Thus, \toolname's witnesses can be viewed as \emph{counter-examples},
thereby connecting it to work on using test generation to find
counter-examples prior to starting a proof~\cite{ACL2Testing, Seidel15}.

@@

Flanagan \etal~\cite{Flanagan96} describe a static debugger for Scheme, which helps
the programmer interactively visualize problematic source-sink flows
corresponding to soft-typing errors. The debugger allows the user to explore
an abstract reduction graph computed from a static value set analysis of
the program. In contrast, \toolname generates witnesses and allows the user
to explore the resulting dynamic execution.
%
Perera \etal~\cite{Perera2012-dy} present a tracing semantics
for functional programs that tags values with their provenance, enabling
a form of backwards program slicing from a final value to the sequence
of reductions that produced it. Notably, they allow the user to supply a
\emph{partial value} --- containing holes --- and present a partial slice,
containing only those steps that affected the the partial value.
% This
% system is designed to answer questions of the form ``Where did this
% value come from?'' and thus is focused on backward exploration.
Perera~\etal\ focus on backward exploration; in contrast, our
visualization supports forward \emph{and} backward exploration, though
our backward steps are more limited.
%
Specifically, we do not support selecting a value and inserting the
intermediate terms that preceded it while ignoring unrelated computation
steps. %; this would be interesting future work.
